---
title: "Degree days to maturation for insects"
author: "Philipp Neubauer"
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    keep_tex: true
    latex_engine: xelatex
    includes:
      in_header: header.tex 
---

```{r echo=F,message=F,warning=FALSE,results='hide'}
require(TaxMeta)
require(knitr)
require(lme4)

opts_chunk$set(cache=TRUE, autodep=TRUE, warning=FALSE ,message=FALSE)

```



# Taxonomic predictions of degree days to maturation for insects

## Load data

Data is from Jarosik et al, Journal of Entomology - DOI: http://dx.doi.org/10.1603/EC11247.

```{r load insects data}
data('insects')
head(insects)
```  

```?insects``` gives info about the data. First, to apply the hierarchical model, we need to set the taxonomy.

```{r get taxonomy, warning=FALSE, message=FALSE}

taxonomy = c('Species','Genus','Family','Order')

```
## Run taxonomic meta-analysis in LME4

After some data manipulation, we can now apply the meta-analysis:

```{r run anlysis, cache=TRUE}


stdise <- function(x) (x-mean(x, na.rm=T))/sd(x, na.rm=T)

insects$lDD <- log10(insects$EADDC)
insects$lBDT <- stdise(insects$BDT)


insects_ss <- insects %>% filter(!is.na(BDT), !is.na(lDD))

lm.ins <- lm(lDD~lBDT+Species,data=insects_ss)
AIC(lm.ins)

lm2.ins <- lm(lDD~lBDT+Species+Genus+Family+Order,data=insects_ss)
AIC(lm2.ins)

lmm.ins <- lmer(lDD~lBDT+(1|Species)+(1|Genus)+(1|Family)+(1|Order),data=insects_ss)
AIC(lmm.ins)

plot(fitted(lm.ins),insects_ss$lDD[!is.na(insects_ss$lDD)])
abline(a=0,b=1)

confint.merMod(lmm.ins,method='Wald')
confint.merMod(lmm.ins)


# reproduce results from 
vcv <- as.data.frame(lme4::VarCorr(lmm.ins))
vcv$vcov/sum(vcv$vcov)
```

These results match those in Jarosik et al. 2011; with most of the variation at the order level (even when including the predictor...). To compare models, I run a 10-fold cross validation on the data.

```{r fig.cap='Observed versus preducted value of $log_{10}$ transformed degree days to maturity in insects, estimated from a taxonomic linear mixed model using the LME4 package.'} 
# out of sample prediction - species
spec <- as.numeric(as.factor(insects_ss$Species))
ssamp <- sample(1:10,max(spec),r=T)
lm.cv.spred = data.frame(matrix(NA,length(spec),3))
for (i in 1:10){
  
  this.lm.ins <- lmer(lDD~lBDT+(1|Species)+(1|Genus)+(1|Family)+(1|Order),data=insects_ss[!spec%in%which(ssamp==i),])
  lm.cv.spred[spec%in%which(ssamp==i),] <-  merTools::predictInterval(this.lm.ins, insects_ss[spec%in%which(ssamp==i),], n.sims=1000, type = "linear.prediction")
  
}
```

## How does TaxMeta compare?

First, a small illustration of differences in estimates for studies without degree day data in the database. Note that for some species in the database, multiple studies are available, only some of which may have DD data. For other data with missing DD data, there may be no other studies for that species - and predictions are thus made from higher taxonomic level information.

\newpage
\blandscape

```{r fig.cap='Predictions from taxonomic models estimated in LME4 and TaxMeta for data with missing degree day information.', fig.width=11,fig.height=8}

insects_res_full <- TaxMeta(insects[!is.na(insects$lBDT),],
                       'lDD',
                       distribution = 'norm',
                       taxonomy = taxonomy,
                       save_tax=taxonomy,
                       type='full',
                       cont_cov = 'lBDT',
                       loo_waic = T,
                       n.chains = 3,
                       n.iter = 50e3,
                       n.thin=25,
                       n.burnin=5e3,
                       return_MCMC=T)

# data without degree day information
ii <- insects[is.na(insects$lDD)&!is.na(insects$lBDT),]

# prediction
lm.pred <- merTools::predictInterval(lmm.ins,newdata=ii)
reSims <- merTools::REsim(lmm.ins, n.sims = 1000)
#merTools::plotREsim(reSims, stat = 'median', sd = TRUE)
#reSims %>% filter(groupFctr=='Family')

lm.pred$Model <- 'LME4'
lm.pred$ind <- 1:nrow(lm.pred)

allfx <- summary(insects_res_full$result)[[2]]
tax.pred <- data.frame(allfx[grepl('pred',rownames(allfx)),c(3,1,5)])
colnames(tax.pred) <- colnames(lm.pred)[1:3]
tax.pred$Model <- 'TaxMeta'
tax.pred$ind <- 1:nrow(tax.pred)
preds <- rbind(tax.pred,lm.pred)

ggplot(preds) + geom_pointrange(aes(x=ind,
                                    y=fit,
                                    ymin=lwr,
                                    ymax=upr, 
                                    linetype=Model),
                            position=position_dodge(width=0.6))+
   xlab('Prediction') + 
  ylab('Prediction') + 
  theme_classic() 
```

We can also compare estimated effects at a particular taxonomic level. For example, comparing estimates for species within families, we can see clear differences between estimates between models:

\newpage

```{r fig.cap='Estiamtes of species effects within three selected families from taxonomic models estiamted in LME4 and TaxMeta.', fig.width=11,fig.height=8}

Specfx <- reSims %>% filter(groupFctr=='Species') %>% dplyr::select(-term,-median)
Specfx$Model <- 'LME4'
Specfx$Family <- insects_ss$Family[match(Specfx$groupID,insects_ss$Species)]
Specfx$groupID <- factor(Specfx$groupID,levels=Specfx$groupID[order(Specfx$Family)])

allfx <- summary(insects_res_full$result)[[1]]
specfx <- data.frame(allfx[grepl('species_mu',rownames(allfx)),]) %>% dplyr::select(-Time.series.SE,-Naive.SE) %>% filter(levels(as.factor(unique(insects[!is.na(insects$lBDT),'Species']))) %in% insects_ss$Species)
specfx$groupFctr <- 'Species'
specfx$groupID <- Specfx$groupID
colnames(specfx)[1:2] <- tolower(colnames(specfx)[1:2])
specfx$Model <- 'TaxMeta'
specfx$Family <- Specfx$Family
specfx$groupID <- factor(specfx$groupID,levels=specfx$groupID[order(specfx$Family)])
spec.fx <- rbind(specfx,Specfx) %>% group_by(Family) %>% filter(n()>70)



ggplot(spec.fx) + geom_pointrange(aes(x=groupID,
                                    y=mean,
                                    ymin=mean-sd,
                                    ymax=mean+sd,
                                    linetype=Model,col=Family),
                                  position=position_dodge(width=0.4)) + 
  xlab('Species') + 
  ylab('Effect') + 
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=8)) 
```
\elandscape
\newpage

Doing cross validation on TaxMeta takes a while, but is done here in full to compare models and illustrate differences:


```{r fig.cap='Observed vs predicted degree day values from ten-fold cross validation applied to models estiamted in LME4 and TaxMeta.'}
this.cvs <- matrix(NA,nrow(insects_ss), 3)
for (i in 1:10){
  insects_test <- insects_ss
  insects_test[spec%in%which(ssamp==i),'lDD'] <- NA
  insects_res_test <- TaxMeta(insects_test,
                              'lDD',
                              distribution = 'norm',
                              taxonomy = taxonomy,
                              type='full',
                              cont_cov = 'lBDT',
                              loo_waic = F,
                              n.chains = 2,
                              n.iter = 50e3,
                              n.thin=25,
                              n.burnin=12e3,
                              return_MCMC=F)
  means <- insects_res_test$result[[1]][,1]
  q1 <- insects_res_test$result[[2]][,1]
  q3 <- insects_res_test$result[[2]][,5]
  pred <- means[grepl('pred',names(means))]
  q1 <- q1[grepl('pred',names(q1))]
  q3 <- q3[grepl('pred',names(q3))]
  this.cvs[spec%in%which(ssamp==i),] <- cbind(q1,pred,q3)
}

par(mfrow=c(1,2))
# LM predictions
plot(insects_ss$lDD,
     lm.cv.spred[,1], 
     pch=16,
     xlim=c(1,4),
     ylim=c(1,4),
     xlab='Observed',
     ylab='Predicted')
arrows(insects_ss$lDD, lm.cv.spred[,2], insects_ss$lDD, lm.cv.spred[,3], length=0.05, angle=90, code=3)
abline(a=0,b=1)


plot(insects_ss$lDD,
     this.cvs[,2], 
     pch=16,
     xlim=c(1,4),
     ylim=c(1,4),
     xlab='Observed',
     ylab='Predicted')
#points(insects_ss$lDD,lm.cv.spred[,1], col=4,pch=16)
#arrows(insects_ss$lDD, lm.cv.spred[,2], insects_ss$lDD, lm.cv.spred[,3], length=0.05, angle=90, code=3,col=4)
arrows(insects_ss$lDD, this.cvs[,1], insects_ss$lDD, this.cvs[,3], length=0.05, angle=90, code=3)
abline(a=0,b=1)
```

Comparing the likelihood of the data under the predictions for the datapoints suggests far better estimates from the TaxMeta model (i.e., greater likelihood).

```{r}
sum(dnorm(insects_ss$lDD,lm.cv.spred[,1],(lm.cv.spred[,3]-lm.cv.spred[,2])/4,log = T))

sum(dnorm(insects_ss$lDD,this.cvs[,2],(this.cvs[,3]-this.cvs[,1])/4,log = T))

```

## Alternative models
We can also compare alternative model formulations in TaxMeta:

```{r run alternative models, cache=TRUE}

insects_res_tax <- TaxMeta(insects_ss,
                       'lDD',
                       distribution = 'norm',
                       taxonomy = taxonomy,
                       save_tax=taxonomy,
                       type='full',
                       cont_cov = 'lBDT',
                       loo_waic = T,
                       n.chains = 3,
                       n.iter = 150e3,
                       n.thin=50,
                       n.burnin=10e3,
                       return_MCMC=F)

insects_res_fixed <- TaxMeta(insects_ss,
                       'lDD',
                       distribution = 'norm',
                       taxonomy = taxonomy,
                       type='fixed',
                       cont_cov = 'BDT',
                       scale=0.01,
                       loo_waic = T,
                       n.chains = 3,
                       n.iter = 150e3,
                       n.thin=50,
                       n.burnin=10e3,
                       return_MCMC=F)

insects_res_gamma <- TaxMeta(insects_ss,
                       'lDD',
                       distribution = 'norm',
                       taxonomy = taxonomy,
                       type='gamma',
                       cont_cov = 'BDT',
                       loo_waic = T,
                       n.chains = 3,
                       n.iter = 150e3,
                       n.thin=50,
                       n.burnin=10e3,
                       return_MCMC=F)

insects_res_unif <- TaxMeta(insects_ss,
                       'lDD',
                       distribution = 'norm',
                       taxonomy = taxonomy,
                       type='uniform',
                       cont_cov = 'BDT',
                       loo_waic = T,
                       n.chains = 3,
                       n.iter = 150e3,
                       n.thin=50,
                       n.burnin=10e3,
                       return_MCMC=F)

```

And compare the results:

```{r compare}

loo_comp <- loo::compare(insects_res_fixed$loo,
             insects_res_gamma$loo,
             insects_res_unif$loo,
             insects_res_tax$loo)

loo_comp <- data.frame(loo_comp)
loo_comp$model <- do.call('rbind',strsplit(rownames(loo_comp),split = '\\$'))[,1]
loo_comp$loo_weight = loo_comp$weight

waic_comp <- loo::compare(insects_res_fixed$waic,
             insects_res_gamma$waic,
             insects_res_unif$waic,
             insects_res_tax$waic)

waic_comp <- data.frame(waic_comp)
waic_comp$model <- do.call('rbind',strsplit(rownames(waic_comp),split = '\\$'))[,1]
waic_comp$waic_weight = waic_comp$weight

comp_tab <- dplyr::inner_join(loo_comp,waic_comp,by='model')

knitr::kable(comp_tab[,c('model','looic','se_looic','loo_weight','waic','se_waic','waic_weight')],
             row.names=F,
             digits=c(0,1,1,2,1,1,2),
             align=c('l','r','r','r','r','r','r'))

```

This comparison suggests that the taxonomic model is the most appropriate for this dataset. A comparison shows how the perception of taxonomic influence on the dependent variable changes among models: in the TaxMeta model, more of the variance is attributed to the species level and less to the genus level relative to the less flexible models.

```{r compare plot, fig.width = 7,  fig.height = 5}

plot(Full=insects_res_tax, 
     Fixed =insects_res_fixed, 
     Gamma =insects_res_gamma,
     Uniform=insects_res_unif)

```
  
